---
title: "Homework 1: USArrests Dataset Analysis"
subtitle: "Advanced Data Programming with R"
author: "Anekha Kallath (24205324)"
date: today
format:
  html:
    theme: minty
execute:
  echo: false
  warning: false
  message: false
---

## Introduction

The in-built R dataset *`USArrests`* is used for this homework. This dataset contains crime statistics for **50** **US states** from **1973**.

The dataset contains statistics on:

-   **Murder**, **Assault**, and **Rape** (arrest rates per 100,000 people)
-   **UrbanPop** : percentage of urban population

```{r}
summary(USArrests)
```

### Summary:

-   **Murder** rates range from **0.8** to **17.4**, with a mean of **7.79**
-   **Assault** is the most prevalent crime, with values ranging from **45** to **337**, and a high mean of **170.8**
-   **UrbanPop**, the percentage of people living in urban areas, ranges from **32%** to **91%**, with a median of **66%**
-   **Rape** arrests range from **7.3** to **46.0**, with a mean of **21.23**

These statistics suggest that **assault** is the dominant form of violent crime in the dataset, while urbanization levels vary considerably across the 50 states.

## Grouping States by Urban Population

```{r}
# Create a categorical variable based on Urban Population - second change - edited comment
USArrests$UrbanGroup <- cut(USArrests$UrbanPop,
                            breaks = c(0, 50, 75, 100),
                            labels = c("Low", "Medium", "High"))
```

```{r}
# Check how many states fall into each group
table(USArrests$UrbanGroup)
```

-   **Low**: ***9*** states with low urban population
-   **Medium**: ***27*** states with moderate urban population
-   **High**: ***14*** states with high urban population

## Urban Population vs Rape Arrests Plot

The urban population categorised into the three groups previously is plotted against the arrests registered for rape cases.

```{r}
# Boxplot of rape arrest rates grouped by urban population categories
boxplot(Rape ~ UrbanGroup, data = USArrests,
        main = "Rape Arrests by Urbanization Level",
        xlab = "Urbanization Level",
        ylab = "Rape Arrests per 100,000",
        col = c("red", "beige", "maroon"),#first change
        border = "gray40")
```

### Conclusion:

The boxplot compares **rape arrest rates** (per 100,000 residents) across three levels of urbanization:

-   **Low Urbanization** states tend to have the lowest rape arrest rates with a tighter spread and a lower median.

-   **Medium Urbanization** states show a wider range of rape arrest rates, including one or two higher outliers.

-   **High Urbanization** states generally have **higher median rape arrest rates** and a broader spread, indicating greater variability and higher extremes.

There is a visible trend where **rape arrest rates increase** with **higher urban population** levels. This pattern may reflect increased reporting, better law enforcement presence in urban areas, or actual higher crime incidence.
